---
title: "Seagrass Fish and Inverts"
author: "ZL Monteith"
date: '2020-07-10'
output: html_document
---

# Seagrass Fish and Inverts Data Package
## Purpose
This script serves to collect, aggregate, and QC fish and macroinvertebrate
survey data from the Hakai Nearshore Seagrass project.


## Setup - packages
This script will use multiple data packages, and source scripts. If you do not
already have the necessary data packages installed, please install them at this
point using `install.packages("PACKAGE_NAME")` in the console window.
```{r setup, include = FALSE, echo = FALSE}
knitr::opts_chunk$set(echo = TRUE)

# load packages
library(tidyverse)
library(magrittr)
library(lubridate)

# load data
fish <- read.csv("../raw-data/seagrass_fish.csv")
inverts <- read.csv("../raw-data/seagrass_inverts.csv")
events <- read.csv("../raw-data/seagrass_events.csv")
```

## Data Quality Check

###Cross Reference Events
In order to ensure completeness of the fish and invertebrate datasets (i.e. all
site visits are included), the data will be aggregated to the event level, and
then cross referenced with the events data also loaded from the EIMS Portal. If
all data are present, the fish and invertebrate events should match the Events
table.

For each event (i.e.. transect - site_id - level visit), there should be two
fish surveys, and two invertebrate surveys. 

```{r event-qc}
# aggregate fish to event level
fish_events <-
  fish %>%
  group_by(date, survey, site_id, collector, collected_start) %>%
  summarise(n_species_obs = n()) %>%
  ungroup() %>%
  group_by(date, survey, site_id) %>%
  summarise(n_fish_surveys = n())

# aggregate inverts to event level
inverts_events <-
  inverts %>%
  group_by(date, survey, site_id, collector, collected_start) %>%
  summarise(n_species_obs = n()) %>%
  ungroup() %>%
  group_by(date, survey, site_id) %>%
  summarise(n_inverts_surveys = n())

# clean and aggregate events
events_count <-
  events %>%
  group_by(date, survey, site_id) %>%
  summarise(n_events = n())  # 

# join fish events with events; there should be two fish surveys for each event
fish_events_check <-
  right_join(fish_events, events_count) %>%
   # remove typo Koeye events still stuck in data portal
  filter(!grepl("KOEYE_ESTUARY_ESTUARY", survey),
         !grepl("KOEYE_ESTUARY ESTUARY", survey),
         !grepl("^KOEYE_KOEYE.*", site_id))

# join invert events with events; there should be two invert surveys for each event
inverts_events_check <-
  right_join(inverts_events, events_count) %>%
   # remove typo Koeye events still stuck in data portal
  filter(!grepl("KOEYE_ESTUARY_ESTUARY", survey),
         !grepl("KOEYE_ESTUARY ESTUARY", survey),
         !grepl("^KOEYE_KOEYE.*", site_id))

# subset out events without the appropriate number of fish surveys
## all data missing (both surveys)
fish_2surveys_missing <-
  fish_events_check %>%
  filter(is.na(n_fish_surveys))

## one survey missing
fish_1survey_missing <-
  fish_events_check %>%
  filter(n_fish_surveys == 1)

## too many surveys; possibly entered twice, or mislabeled
fish_too_many_surveys <-
  fish_events_check %>%
  filter(n_fish_surveys > 2)

## join all fish events needing QC
fish_events_to_QC <-
  rbind(fish_2surveys_missing,
        fish_1survey_missing,
        fish_too_many_surveys) %>%
  mutate(issue = case_when(is.na(n_fish_surveys) ~ "both surveys missing",
                           n_fish_surveys == 1 ~ "one survey missing",
                           n_fish_surveys > 2 ~ "too many surveys")) %>%
## 2015 data not in the data portal currently
  filter(year(date) != 2015) %>%
  
## For 2018 onward, no fish and invert surveys conducted for Goose SE, McMullin, and Triquet
  filter(!(year(date) >= 2018 & survey == "GOOSE_SOUTHEAST"), 
           !(year(date) >= 2018 & survey == "MCMULLIN_NORTH"), 
             !(year(date) >= 2018 & survey == "MCMULLIN_SOUTH"), 
               !(year(date) >= 2018 & survey =="TRIQUET_NORTH"), 
                 !(year(date) >= 2018 & survey =="TRIQUET_BAY"))

## save `fish_events_to_QC` to .csv:
write.csv(fish_events_to_QC, "../raw-data/fish_events_to_QC.csv")

# subset out events without the appropriate number of invert surveys
## all data missing (both surveys)
inverts_2surveys_missing <-
  inverts_events_check %>%
  filter(is.na(n_inverts_surveys))

## one survey missing
inverts_1survey_missing <-
  inverts_events_check %>%
  filter(n_inverts_surveys == 1)

## too many surveys; possibly entered twice, or mislabeled
inverts_too_many_surveys <-
  inverts_events_check %>%
  filter(n_inverts_surveys > 2)

## join all invert events needing QC
inverts_events_to_QC <-
  rbind(inverts_2surveys_missing,
        inverts_1survey_missing,
        inverts_too_many_surveys) %>%
  mutate(issue = case_when(is.na(n_inverts_surveys) ~ "both surveys missing",
                           n_inverts_surveys == 1 ~ "one survey missing",
                           n_inverts_surveys > 2 ~ "too many surveys")) %>%
  
## 2015 data not in the data portal currently
  filter(year(date) != 2015) %>%

## For 2018 onward, no fish and invert surveys conducted for Goose SE, McMullin, and Triquet
  filter(!(year(date) >= 2018 & survey == "GOOSE_SOUTHEAST"), 
           !(year(date) >= 2018 & survey == "MCMULLIN_NORTH"), 
             !(year(date) >= 2018 & survey == "MCMULLIN_SOUTH"), 
               !(year(date) >= 2018 & survey =="TRIQUET_NORTH"), 
                 !(year(date) >= 2018 & survey =="TRIQUET_BAY"))

## save `fish_events_to_QC` to .csv:
write.csv(inverts_events_to_QC, "../raw-data/inverts_events_to_QC.csv")
```


### Checking for suspicious values

Fix names so there aren't multiple names for single species
```{r}
# import list of acceptable fish names
fishnames_accepted <- read.csv("../raw-data/fish_names_list.csv", header = FALSE)
colnames(fishnames_accepted) <- "name"

# create list of fish names in data
fishnames_data <- unique(fish$species) %>% sort %>% data.frame
colnames(fishnames_data) <- "name"

# filter out names in data that differ from acceptable naming convention
anti_join(fishnames_data, fishnames_accepted)

# fix names that don't match acceptable naming convention
fishnames_data$name[c(1, 2, 3, 4, 5, 7, 8, 9, 10, 12, 15, 17, 26, 27, 28, 29, 34, 36, 37, 38, 48, 49, 59, 60, 61, 62, 64, 65, 66, 67, 68, 69, 73, 74, 75, 76, 77, 78)] <- c("flounder/arrow_tooth", "goby/bay", "pipefish/bay", "skate/big", "sole/c-o", "rockfish/deacon", "sole/english", "sole/english", "sole/c-o", "flatfish/soledab", "goby/bay", "greenling/white-spotted", "flounder/pacific_sanddab", "flounder/pacific_sanddab", "greenling/painted", "flounder/pacific_sanddab", "prickleback/snake", "sole/rock", "rockfish/deacon", "rockfish/deacon", "sole/sand", "flounder/pacific_sanddab", "skate/unknown", "sole/unknown", "sole/english", "sole/unknown", "flatfish/soledab", "threespine_stickleback", "threespine_stickleback", "threespine_stickleback", "threespine_stickleback", "threespine_stickleback", "cod/unknown", "flatfish/unknown", "flatfish/unknown", "rockfish/vermillion", "greenling/white-spotted", "greenling/white-spotted")

# check that all names match ones in acceptable names list now
anti_join(fishnames_data, fishnames_accepted)

```

Check fish lengths and abundances for suspicious values, using Q-Q plots to look at data values vs. quantiles:
```{r}
#create qqplot
fishlengths_qqplot <- fish %>%
  ggplot(aes(col = species)) + 
  stat_qq(aes(sample = length_cm)) + 
  stat_qq_line(aes(sample = length_cm)) + 
  facet_wrap( ~ species) + 
  theme_classic() + 
  theme(legend.position = "none")

#take a closer look at qqplots that look a little fishy (hah)
#code to pull up smaller subset of qqplots 

#filter out suspicious-looking lengths
# fishQC_length_cm <- fish %>%
#   filter(species == "sole/english" & length_cm > 25 |
#            species == "gunnel/crescent" & length_cm < 25)

```
